Quantifying the home ground advantage

### 2022-04-20 inst/home_advantage.Rmd
update home advantage work

```{r,label=makehyper2}
H <- hyper2(pnames=c(teams,"home"))
```

First include the home games won:

```{r,label=includehome}
for(i in seq_len(nrow(home_games_won))){
  for(j in seq_len(ncol(home_games_won))){
    if(i != j){  # diagonal means a team plays itself, meaningless
      game_winner <- teams[i]
      game_loser  <- teams[j]
      no_of_matches <- home_games_won[i,j] # won with home strength helping
      H[c(game_winner,           "home")] %<>% inc(no_of_matches)  # won with home strength helping
      H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)
    }
  }
}
```

Now away games won:

```{r,label=awaygameswon}
for(i in seq_len(nrow(away_games_won))){
  for(j in seq_len(ncol(away_games_won))){
    if(i != j){
      game_winner <- teams[i]
      game_loser  <- teams[j]
      no_of_matches <- away_games_won[i,j] # won without the benefit of home strength
      H[c(game_winner                  )] %<>% inc(no_of_matches)  # won without home ground helping...
      H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)  # home strength nevertheless on denominator
    }
  }
}
```

Thus

```{r,label=showH}
H
```

Visually:

@@

Now a likelihood ratio for the hypothesis that home strength (or home
advantage) is zero.  First define a function `f()` that maximises the
likelihood subject to the home strength being *less than* its
argument:

```{r,label=nullofzerohomestrength}
f <- function(home){
	small <- 1e-3
	maxp(H,startp=rep((1-small)/4,4),give=TRUE,fcm=t(rep(1,4)),fcv=1-home)$value
}
```

See how the `fcm` and `fcv` constraints make the sum of a,b,c,d,
greater than `1-home`.  Thus 'home' [which is the fillup value,
$p_\mathrm{home}=1-(p_A+p_B+p_C+p_D)$] is constrained to be less than
argument `home`.  Using this function, the likelihood ratio test is
straightforward:

```{r,label=showLambda,cache=TRUE}
Lambda <- -2*(f(0.01)-maxp(H,give=TRUE)$value)
Lambda
```

Under the null, Wilks's theorem gives us that Lambda is distributed as
a chi-square with 1 degree of freedom (clearly rejected).  However, it
is always interesting to plot likelihood functions.  To plot a profile
likelihood we can constrain the home to be greater than a particular
value:

```{r,calcprofilelikelihood}
g <- function(home){
	small <- 1e-3
	maxp(H,startp=rep(small/4,4),give=TRUE,fcm=-t(rep(1,4)),fcv=home-1)$value
}
```

```{r,showprofilelikelihood,cache=TRUE}  # this one takes the time
home_strength <- seq(from=0.2,to=0.6,length=30)
support <- pmin(sapply(home_strength,f),sapply(home_strength,g))
plot(home_strength,support-max(support),
     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
abline(h=c(0,-2))
```

>>>>

```{r,label=makehyper2,cache=TRUE}
H <- home_away(home_games_won,away_games_won)
H
options("use_alabama" = TRUE) # to avoid the stupid wmmin bug
specificp.gt.test(H,"home",0)
```

We see strong evidence to support the contention that home advantage is real.
Visually:

@@

A profile likelihood diagram:

```{r,label=showprofilelikelihood, cache=TRUE}
home_strength <- seq(from=0.2,to=0.6,length=20)
plot(home_strength,profsupp(H,"home",home_strength),type="b",pch=16,
     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
abline(h=c(0,-2))
```
